\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2020-03-25 k-planar.tex
Relating our results to  Pilipczuk and Siebertz

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centered colouring  \citep{micek:personal},
% and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}.
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth.
% Say that a graph $G$ admits a $(t,\ell)$-partition if $V(G)$ has a partition $\mathcal{P}$ of layered width at most $\ell$ and the treewidth of $H:=G/\mathcal{P}$ is at most $t$.
They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactenum}[(a)]
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactenum}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions and corresponding layerings.
%Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define.
%
% \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.
%
% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:
The same authors proved the following generalisation of \cref{PlanarBasic,PlanarPartition} for graphs embeddable on other surfaces.\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.}

@@

@@

\begin{proof}
  We count the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap L_i$.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Centered Colourings}

@@

@@

The following lemma is implicitly due to \citet{micek:personal}. We include the proof for completeness.

@@

\begin{lem}[\citep{micek:personal}]
\label{p-centered}
Every graph $G$ that has an $H$-partition of layered width at most $\ell$ has $\chi_p(G)\le \ell (p+1)\, \chi_p(H)$.
\end{lem}

>>>>

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centered colouring  \citep{debski.felsner.ea:improved},
% and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}.
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth.
% Say that a graph $G$ admits a $(t,\ell)$-partition if $V(G)$ has a partition $\mathcal{P}$ of layered width at most $\ell$ and the treewidth of $H:=G/\mathcal{P}$ is at most $t$.
Dujmović \etal\ define a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactenum}[(a)]
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and \label{subthm:eight-one}
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.\label{subthm:three-tree}
\end{compactenum}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions and corresponding layerings.
%Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define.
%
% \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.
%
% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:
The same authors proved the following generalisation of \cref{PlanarBasic,PlanarPartition} for graphs embeddable on other surfaces.\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.}

@@

At the core of their work is an elegant proof of the following result of  \citet{pilipczuk.siebertz:polynomial-soda} that was introduced in the context of centered colourings (see \cref{sec:centered-colourings}):
\begin{thm}[\citep{pilipczuk.siebertz:polynomial-soda}]\thmlabel{ps}
  Every planar triangulation $G$ has an $H$-partition $\mathcal{P}$
  such that $H$ has treewidth at most 8 and, for each $P\in\mathcal{P}$, $G[P]$ is a shortest path in $G$.
\end{thm}

Indeed, \cref{PlanarPartition}.\ref{subthm:eight-one} is a slight strengthening of \cref{thm:ps} in which $G$ has a distinguished vertex $r$ and, for each $P\in\mathcal{P}$ no two vertices of $P$ have the same distance to $r$.

@@

\begin{proof}
  Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.  We say that a vertex $w\in Y_x$ \emph{contributes} a vertex $v\in S_x$ if $v$ participates in $x$.
  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes at most one vertex, itself, to $S_x\cap L_i$.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Centered Colourings}
\seclabel{centered-colourings}

@@

\citet{pilipczuk.siebertz:polynomial-soda} and \citet{debski.felsner.ea:improved} use \cref{thm:ps} and \cref{PlanarPartition}.\ref{subthm:three-tree}

@@

The following lemma is implicitly in the work of \citet[Proof of Theorem 2.1]{debski.felsner.ea:improved}. We include the proof for completeness.

@@

\begin{lem}[\citep{debski.felsner.ea:improved}]
\label{p-centered}
Every graph $G$ that has an $H$-partition of layered width at most $\ell$ has $\chi_p(G)\le \ell (p+1)\, \chi_p(H)$.
\end{lem}
